### 2025-06-03 writing/manuscript.tex
draft 2

\pagebreak
\section{Introduction}
Equally qualified individuals may face very different labor market outcomes depending on their socioeconomic status (SES). A key driver of this inequality is due to differences in social capital,\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} with recent empirical work characterizing its most important facet as the “share of high-SES friends among individuals with low-SES” because it correlates strongly with labor market income \citep{chetty2022social_a}. A lack of social capital means a lack of access to individuals with influential (higher paid) jobs and job opportunities. In economic terms, it implies having worse outcomes when using one's network to find jobs conditional on the capacity on leveraging one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use. \textcolor{red}{fill in citaiton sociology reading from slides}}

Referral hiring, the formal or informal process where firms ask workers to recommend qualified candidates for job opportunities, is a common labor market practice which makes evident the role of differences in social capital. As referrals must orginate from the networks of referrers, the composition of referrer networks becomes a crucial channel that may propagate inequality: Similar individuals across socio-demographic characteristics tend to form connections at higher rates \citep{mcpherson2001birds}, making across SES (low-to-high) connections less likely than same-SES connections \citep{chetty2022social_a}. Referrals will thus reflect similarities in socio-demographic characteristics present in networks even in the absence of biases in the referral procedure, i.e., referring at random from one's network according to some productivity criteria.

Evidence shows referrals can be biased under substantial pay-for-performance incentives beyond what is attribuable to differences in network compositions, at least for the case of gender \citep{beaman_job_2018,hederos2025gender}. A similar bias against low-SES may further exacerbate outcomes of low-SES individuals: If job information are in the hands of a select few high-SES which low-SES have already limited network access to (social capital hypothesis), and high-SES referrers are biased against low-SES, referring other high-SES at higher rates than their network composition, we expect referral hiring to disadvantagee low-SES. The empirical question we answer is whether there is a bias against low-SES once we account for the network SES compositon in a controlled setting.

In this study, we focus on the role of SES in referrals by investigating whether individuals who are asked to refer someone tend to refer a same-SES candidate. We also explore potential mechanisms behind referral patterns under different incentives. To this end, we conducted a lab-in-the-field experiment with 734 participants in a Colombian university where different SES groups mix together. Participants were instructed to refer a qualified student for tasks similar to the math and reading parts of the national university entry exam (equivalent of SAT in US system). To incentivize participants to refer qualified candidates, we set earnings dependent on referred candidates' actual university entry exam scores.

Referral hiring in the labor market can range from firm-level formal referral programs asking employees to bring candidates to simply passing on job opportunities between network members \citep{topa2019social}. As participants in our study are students at the university referring based on exam scores, we abstract away from formal referral programs with defined job openings. Our setting instead resembles situations where contacts share opportunities with each other without the need for the referred candidate to take any action. This also eliminates reputational concerns as there is no firm (see for example \citet{witte2021workers,bandiera2009social}). At the same time, national university entry exam scores are still objective, widely accepted measures of ability, and we show evidence that referrers in our setting possess accurate information about these signals.

We find strong evidence that networks of high- and low-SES participants exhibit SES-homophily. Both groups are connected at higher rates with their own SES group than what would be at random given actual group shares at the university (Figure \ref{availability_lses}). As students take more courses together within the same program, their networks dwindle in size (Figures \ref{share_same_program_by_tie} and \ref{conns_by_tie}), and become more homogenous in SES-shares (Figure \ref{share_same_ses_by_tie}). We identify selection into academic programs as a key mechanism. The private university where our study takes place implements exogenous cost-based program pricing and does not offer SES-based price reductions. These result in programs with very large cost differences within the same university (Figure \ref{fees_with_boxplot}). We find that average yearly fees paid increase with SES, and the high-SES share in the most expensive program at the university, medicine, drives the network segregation across SES (Figure \ref{ses_distribution_by_fees}).

@@

>>>>

\pagebreak
\section{Introduction}
Equally qualified individuals may face very different labor market outcomes depending on their socioeconomic status \citep{stansbury2024class}. A key driver of this inequality is due to differences in social capital,\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} with recent empirical work characterizing its most important facet as the “share of high-SES friends among individuals with low-SES” as it correlates strongly with labor market income \citep{chetty2022na}. A lack of social capital means a lack of access to individuals with influential (higher paid) jobs and job opportunities. In economic terms, it implies having worse outcomes when using one's network to find jobs conditional on the capacity on leveraging one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use. \textcolor{red}{fill in citaiton sociology reading from slides}}

Referral hiring, the formal or informal process where firms ask workers to recommend qualified candidates for job opportunities, is a common labor market practice which makes evident the role of differences in social capital. As referrals must orginate from the networks of referrers, the composition of referrer networks becomes a crucial channel that may propagate inequality: Similar individuals across socio-demographic characteristics tend to form connections at higher rates \citep{mcpherson2001birds}, making across SES (low-to-high) connections less likely than same-SES connections \citep{chetty2022na}. Referrals will thus reflect similarities in socio-demographic characteristics present in networks even in the absence of biases in the referral procedure, i.e., referring at random from one's network according to some productivity criteria.

Yet, experimental evidence shows referrals can be biased under substantial pay-for-performance incentives beyond what is attribuable to differences in network compositions, at least for the case of gender \citep{beaman_job_2018,hederos2025gender}. A similar bias against low-SES may further exacerbate outcomes of low-SES individuals: If job information are in the hands of a select few high-SES which low-SES have already limited network access to (social capital hypothesis), and high-SES referrers are biased against low-SES, referring other high-SES at higher rates than their network composition, we should expect referral hiring to further disadvantagee low-SES. The empirical question we answer is whether there is a bias against low-SES once we account for the network SES compositon in a controlled setting.

In this study, we study inequalities related to SES combining a university-wide cross-sectional network data set comprising over 4,500 students in which classroom interactions are recorded along with individual attributes. We focus on the role of SES in referrals by experimentally investigating whether individuals who are asked to refer a peer tend to refer a same-SES candidate. We also explore potential mechanisms behind referral patterns by randomizing participants into two different incentive structures. To this end, we conducted a lab-in-the-field experiment with 734 students in a Colombian university. Participants were instructed to refer a qualified student for tasks similar to the math and reading parts of the national university entry exam (equivalent of SAT in US system). To incentivize participants to refer qualified candidates, we set earnings dependent on referred candidates' actual university entry exam scores.

Referral hiring in the labor market can range from firm-level formal referral programs asking employees to bring candidates to simply passing on job opportunities between network members \citep{topa2019social}. As our participants are students at the university and refer based on exam scores, we abstract away from formal referral programs with defined job openings. Our setting instead resembles situations where contacts share opportunities with each other without the need for the referred candidate to take any action. This eliminates reputational concerns as there is no firm (see for example \citet{witte2021workers,bandiera2009social}). At the same time, national university entry exam scores are still objective, widely accepted measures of ability, and we show evidence that referrers in our setting possess accurate information about these signals.

In a university setting, class attendance provides essential opportunities for face-to-face interaction between students. On the one hand, this reduces network segregation by providing ample opportunities to meet across-SES, because of the exposure to an equal or higher level of high-SES compared to the population \citep{chetty2022n}. On the other hand, as students take more and more classes together, their similarities across all observable characteristics tend to increase \citep{kossinets_origins_2009}, which should drive the high- and low-SES networks to segregate. Our setting is ideal to study these opposing forces: First, The very high level of income inequality and existence of deeply rooted historical groups in Colombia makes SES differences extremely visible in access to tertiary education, where the rich and poor typically select into different institutions \citep{jaramillo-echeverri2023}. Yet, thanks to the particular standing of the institution we have chosen for this study (Figure \ref{school_quality_income}), different SES groups including both low- and high-SES mix together in this university.

We find strong evidence that networks of high- and low-SES participants exhibit same-SES bias. Both groups are connected at higher rates with their own SES group than what would be at random given actual group shares at the university (Figure \ref{availability_lses}). As students take more courses together within the same program, their networks dwindle in size (Figures \ref{share_same_program_by_tie} and \ref{conns_by_tie}), and become more homogenous in SES-shares (Figure \ref{share_same_ses_by_tie}). We identify selection into academic programs as a key mechanism. The private university where our study took place implements exogenous cost-based program pricing and does not offer SES-based price reductions. These result in programs with very large cost differences within the same university (Figure \ref{fees_with_boxplot}). We find that average yearly fee paid per student increases with SES, and the high-SES share in the most expensive program at the university, medicine, drives the network segregation across SES (Figure \ref{ses_distribution_by_fees}).

@@

\begin{figure}[H]
  \centering
  \caption{Participant network size and tie strength by time spent at UNAB}\label{school_quality_income}
  \includegraphics[width=0.6\linewidth]{../figures/school_quality_income.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure displays the average number of connections for referrers in blue and the average number of classes they have taken together with their connections in green across semesters spent at UNAB. The data shows an increase in the number of classes taken together as students progress in their programs, with the connections peaking around 7 semesters and dropping as certain students finish their bachelor's. Figure reproduced from \citep{fergusson2021distincion}
  \end{tablenotes}
\end{figure}
